## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <= 5) {
  print(a)
} else {
  print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5) {
  print("small")
} else if (a > 5 & a <= 10) {
  print("medium")
} else if (a > 10 | a == 7) {
  print("large")
} else {
  print("huge")
}
    
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5

if (a / 2 != 0) {
  if (a == 5){
    print("a is 5")
  }
} 
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) {
  # Paste is a concatenation function
  print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

# Change the first, third, and sixth elements to the value 6
for (i in c(1,3,6)){
  a_vector[i] <- 6
}

a_vector


```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10) {
  # increase b by 2
  b <- b + 2
  # print b
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
# Start with a vector of temperatures
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

# Then we want a function thatconverts from C to F:
c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) + 32
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f((test_temp))
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
# 1 is for rows, 2 is for columns
apply(test_mat, 1, sum)
apply(test_mat, 2, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #Given a set of numbers, return the additive inverse of each. Each positive becomes negatives, and the negatives become positives.

    invert <- function(vec){
      invertNum <- vec*-1
      return(invertNum)
    }
    ```

    ```{r}
    #This kata is about multiplying a given number by eight if it is an even number and by nine otherwise.

    simple_multiplication <- function(number) {
      if (number %% 2 == 0) {
      return(number*8)
    } else {
      return(number*9)}
    } 
    ```

    ```{r}
    #Clock shows h hours, m minutes and s seconds after midnight.
    #Your task is to write a function which returns the time since midnight in milliseconds.

    past <- function(h, m, s) {
      timesincemid <- h * 3600000 + m * 60000 + s * 1000
      return(timesincemid)
    }
    ```

    ```{r}
    #Write a function that takes an array of numbers and returns the sum of the numbers. The numbers can be negative or non-integer. If the array does not contain any numbers then you should return 0.

    sum_array <- function(a){
      summedarr <- sum(a)
      return(summedarr)
    }
    ```

    ```{r}
    #Create a function finalGrade, which calculates the final grade of a student depending on two parameters: a grade for the exam and a number of completed projects.
    #This function should take two arguments: exam - grade for exam (from 0 to 100); projects - number of completed projects (from 0 and above);
    #This function should return a number (final grade). There are four types of final grades:
    #100, if a grade for the exam is more than 90 or if a number of completed projects more than 10.
    #90, if a grade for the exam is more than 75 and if a number of completed projects is minimum 5.
    #75, if a grade for the exam is more than 50 and if a number of completed projects is minimum 2.
    #0, in other cases

    final_grade <- function(exam, projects){
      if (exam > 90 | projects > 10) {
      finalgrade <- 100
    } else if (exam > 75 & projects >= 5) {
      finalgrade <-90
    } else if (exam > 50 & projects >= 2) {
      finalgrade <- 75
    } else {
      finalgrade <- 0
      return(finalgrade)
    }
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    Jen_array <- array(
      c(10,9,8,7,6,5,4,3,2,1,11, 12,13,14,15,16,17,18,19,20),
      dim <- c(4,5,1)
    )
    Jen_array

    fizzbuzzfunc <- function(number){
    if (number %% 3 == 0 & number %% 5 == 0) {
      print("fizzbuzz")
    } else if (number %% 3 == 0) {
      print("fizz")
    } else if (number %% 5 == 0) {
      print("buzz")
    } else { print("None")
    }  
    }
    #print(fizzbuzzfunc(9))


    apply(Jen_array, c(1,2), fizzbuzzfunc)    

    #create an array of 1-20 and map over (extra pts)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    marioStairs <- function(stairs){
      stair <- 0
      while (stair < stairs){
        stair <- stair + 1
        print(strrep('X', stair))
        #print("X") *stair
    }
    }

    print(marioStairs(4))



    ```
